Faithful short-read mapping with Sesame

### 2019-09-17 manuscript.tex
First draft of the introduction

\title{Faithful short read mapping with Sesame}

\author{%
Ruggero Cortini\,$^{1}$,
Eduard Valera Zorita\,$^{1}$,
Guillaume J. Filion\,$^{1,2,3}$
\footnote{To whom correspondence should be addressed.
Email: guillaume.filion@gmail.com}}

\address{%
$^{1}$Center for Genomic Regulation (CRG), The Barcelona Institute
of Science and Technology, Dr. Aiguader 88, Barcelona 08003, Spain;
$^{2}$University Pompeu Fabra (UPF), Barcelona, Spain;
$^{3}$present address: Department of Biological Sciences, University
of Toronto Scarborough, Toronto, ON, Canada.}
% Affiliation must include:
% Department name, institution name, full road and district address,
% state, Zip or postal code, country

@@

Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text.
Text \cite{1}.

Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
\begin{align*}
&\mathrm{Ascorbate} + \mathrm{EDTA} \cdot \mathrm{Fe}^{3+} \to
\hbox{Oxidized ascorbate}
\\
&\mathrm{EDTA} \cdot \mathrm{Fe}^{2+} + \mathrm{H}_2
\mathrm{O}_2 \to
\mathrm{EDTA} \cdot \mathrm{Fe}^{3+} + \cdot
\mathrm{OH} + \mathrm{OH}^-
\end{align*}
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.

>>>>

\title{Faithful short-read mapping with Sesame}

\author{%
Ruggero Cortini\,$^{1,\text{\#}}$,
Eduard Valera Zorita\,$^{1,\text{\#}}$,
Guillaume J. Filion\,$^{1,2,3}$
\footnote{To whom correspondence should be addressed.
Email: guillaume.filion@gmail.com}}

\address{%
$^{1}$Center for Genomic Regulation (CRG), The Barcelona Institute
of Science and Technology, Dr. Aiguader 88, Barcelona 08003, Spain;
$^{2}$University Pompeu Fabra (UPF), Barcelona, Spain;
$^{3}$present address: Department of Biological Sciences, University
of Toronto Scarborough, Toronto, ON, Canada; $^{\text{\#}}$ equal
contributions.}
% Affiliation must include:
% Department name, institution name, full road and district address,
% state, Zip or postal code, country

@@

High throughput DNA sequencing is now a standard technology in the
academia and in the industry, with countless applications in diagnosis,
forensics, surveillance and research \cite{1}. The Illumina short-read
technology currently dominates the market of DNA sequencing, making the
associated software an important target for optimization.

The standard way to identify a read is to map it to a known reference
sequence, typically a genome. Work on short-read mappers has been
traditionally focused on improving the speed, reducing the memory
footprint and increasing the accuracy. Another important focus has been to
develop variations to address the actual needs of the user, as different
problems often entail different read types (e.g., compare CHiP-seq and
Hi-C).

Meanwhile, progress has been more limited on another aspect of the mapping
process: \emph{faithfulness}. Mapping algorithms are heuristics, so there
is always a chance that the proposed location of a read is wrong.
Faithfulness is the capacity to correctly estimate this probabilty.
Importantly, one can reduce the probability of errors without having to
measure it, so accuracy and faithfulness are usually unrelated.

The importance of faithfulness is recognized in the specifications of the
SAM format \cite{1} that include a \emph{mapping quality} score. The
standard defines it as $-10 \cdot \log_{10}Pr$\{mapping position is
wrong\}, but some mappers use a qualitative scale (e.g., STAR), and
some others do not compute it (e.g., GEM). As for the mappers that comply
with the SAM standard, the scores are undocumented, causing
puzzlment and frustration\footnote{http://www.acgt.me/blog/2015/3/17/more-madness-with-mapq-scores-aka-why-bioinformaticians-hate-poor-and-incomplete-software-documentation}.

Achieving faithful read mapping is a hard problem, for which this is
presently no solution. Yet, this is an important challenge for the
following reasons:
$i$ in many applications it is essential to know the risk associated with
every read (e.g., contaminated material such as ancient DNA);
$ii$ a mapper is easier to use if error rates are what they claim to
be;
$iii$ good calibrations open opportunities to improve the speed-accuracy
tradeoff.

We recently proposed a strategy to compute the error rate of different
seeding strategies used in short-read mappers \cite{1}. This strategy
borrows some basic concepts from analytic combinatorics \cite{1} to
develop a formal computation framework. Here we present a C implementation
of this framework in a library called Sesame. We also develop a
statistical method to estimate the parameters required to compute seeding
probabilities. We demonstrate the feasibility of our approach by
developing a naive mapper based on Sesame. Not only is this mapper
faithful, but it also reveals the existence of ``super reads'' that can be
mapped with much higher confidence than was previously thought possible.

\begin{align*}
&\mathrm{Ascorbate} + \mathrm{EDTA} \cdot \mathrm{Fe}^{3+} \to
\hbox{Oxidized ascorbate}
\\
&\mathrm{EDTA} \cdot \mathrm{Fe}^{2+} + \mathrm{H}_2
\mathrm{O}_2 \to
\mathrm{EDTA} \cdot \mathrm{Fe}^{3+} + \cdot
\mathrm{OH} + \mathrm{OH}^-
\end{align*}
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
